\documentclass{midl} % Include author names

% The following packages will be automatically loaded:
% jmlr, amsmath, amssymb, natbib, graphicx, url, algorithm2e
% ifoddpage, relsize and probably more
\usepackage{booktabs}
\usepackage{siunitx}

% Header for extended abstracts
\jmlrproceedings{MIDL}{Medical Imaging with Deep Learning}
\jmlrpages{}
\jmlryear{2026}

\jmlrworkshop{Short Paper -- MIDL 2026 submission}
\jmlrvolume{-- Under Review}
\editors{Under Review for MIDL 2026}

\title[MRI-derived Face Age vs Brain Age]{MRI-derived Face Age vs Brain Age from the Same T1 Scan: a Proof-of-Concept Pipeline}

\midlauthor{%
  \Name{Ekaterina Kondrateva\nametag{$^{1}$}} \Email{kondratevakate@gmail.com}\\
  \Name{Ramil Khafizov\nametag{$^{1}$}}\\
  \Name{Gleb Bobrovskikh\nametag{$^{1}$}}\\
  \addr $^{1}$ Independent researchers
}

\begin{document}

\maketitle

\begin{abstract}
  We present a proof-of-concept pipeline for extracting both face age and
  brain age from the same non-defaced T1-weighted MRI volume. The face branch
  renders a frontal portrait from the MRI surface and applies an existing
  FaceAge model; the brain branch estimates age with SynthBA on the same
  acquisition. Evaluating on the IXI dataset ($N{=}406$ for face age,
  $N{=}105$ for brain age, $N{=}93$ overlapping), we find that SynthBA
  achieves MAE~$6.33$ years (Pearson $r{=}0.84$) while calibrated FaceAge
  reaches MAE~$10.24$ years ($r{=}0.83$). The two aging gaps computed from
  the same scan show a statistically significant positive correlation
  ($r{=}0.31$, $\rho{=}0.32$, $p{<}0.01$), suggesting partial but
  incomplete overlap between facial and cerebral aging signals. Scanner
  reproducibility on the single-subject SIMON dataset (99 scans, 36 scanners)
  shows SynthBA brain-age predictions are highly stable (SD~$1.21$ years),
  though with a systematic negative bias of $-16.16$ years on this subject.
  The pipeline and all code are openly available.
\end{abstract}

\begin{keywords}
  brain age, face age, MRI, biological age, age gap, reproducibility, IXI, SIMON
\end{keywords}

% -----------------------------------------------------------------------
\section{Introduction}
% -----------------------------------------------------------------------

Standard MRI pre-processing pipelines routinely remove facial structure
(``defacing'') to protect participant identity before data sharing.
However, non-defaced T1-weighted volumes---common in single-site research
and legacy datasets such as IXI~\citep{ixi} and
SIMON~\citep{simon}---contain both a brain and a complete facial surface
within the same acquisition.  This raises a natural question: do the
\emph{facial} morphology and the \emph{brain parenchyma} record the same
aging trajectory, or do they carry partly independent biological information?

Prior work on brain age~\citep{peng2021sfcn, lemaitre2022synthba} and face
age~\citep{bontempi2025faceage} treats these as entirely separate modalities
from different instruments.  No existing study compares the two gap
estimates from a \emph{single scan}.  Such a comparison avoids the
domain shift of photo-vs-MRI studies and allows a controlled test of
whether the face and brain age together or drift apart.

We contribute: (1) an open pipeline that renders MRI-derived face images and
runs both modalities in parallel; (2) a systematic comparison of render
strategies; and (3) an analysis of face-age/brain-age gap correlation and
scanner reproducibility on two public datasets.

% -----------------------------------------------------------------------
\section{Methods}
% -----------------------------------------------------------------------

\paragraph{Face branch.}
Given a T1 NIfTI volume, we extract the iso-surface with marching cubes
(intensity threshold~$t{=}30$) and render a frontal portrait using
PyVista~\citep{pyvista} with nine viewpoints ($\pm 20°$ yaw/pitch) and
randomised lighting.  We apply FaceAge~\citep{bontempi2025faceage}, a
ResNet-50 model trained on 56\,304 curated photographs, to each render and
average the predictions.  We also tested photorealistic augmentation via
stable diffusion to close the photo--render domain gap, but found this
substantially worsened accuracy (see \tableref{tab:face}).

To correct the systematic positive bias of FaceAge on MRI renders, we
fit a linear calibration $\hat{y}_{\text{chron}} = a \cdot \hat{y}_{\text{face}} + b$
on a held-out set of 20 subjects and apply it to the remaining 80.

\paragraph{Brain branch.}
Skull stripping is performed with SynthStrip~\citep{synthstrip}.  We then
run SynthBA~\citep{lemaitre2022synthba}, a segmentation-based model robust
to acquisition variability, on the stripped volume.

\paragraph{Datasets.}
The \textbf{IXI} dataset~\citep{ixi} contains 561 T1 scans of healthy adults
(age 20--86 years) from three London sites.  After manual quality control of
frontal renders (removing scans with floating artefacts or missing facial
structure), 406 subjects were retained for face-age analysis; the brain-age
model was run on a 105-subject subset.  Both sets share 93 subjects, enabling
paired gap analysis.

The \textbf{SIMON} dataset~\citep{simon} provides 99 T1 scans of a single
healthy male (age range 29--46 years) collected across 73 sessions on 36
different scanners, enabling a direct test of scan-rescan reproducibility.

% -----------------------------------------------------------------------
\section{Results}
% -----------------------------------------------------------------------

\paragraph{Face-age accuracy on IXI.}
\tableref{tab:face} compares render strategies.  Multiview averaging with a
higher intensity threshold outperforms single-view rendering.  Photorealistic
augmentation is counterproductive: the generated images shift the appearance
distribution away from the photographs FaceAge was trained on, causing a
near-perfect positive bias equal to the MAE.  Post-hoc linear calibration
(fit on 20 subjects, applied to 80) recovers most accuracy, reaching
MAE~$10.24$ years and reducing bias to $-2.12$ years.

\begin{table}[htbp]
\floatconts{tab:face}
  {\caption{Face-age prediction accuracy on IXI. Calibration formula:
    $\hat{y}_{\text{chron}} = 1.010\,\hat{y}_{\text{face}} - 17.049$.}}
  {\begin{tabular}{lrrr}
    \toprule
    \textbf{Method} & \textbf{MAE (yr)} & \textbf{RMSE (yr)} & \textbf{Bias (yr)} \\
    \midrule
    Multiview renders ($N{=}406$)              & 11.34 & 13.83 & $+10.04$ \\
    Photorealistic renders (SD-augmented)      & 19.91 & 21.12 & $+19.91$ \\
    Multiview + linear calibration             & \textbf{10.24} & --- & $-2.12$ \\
    \bottomrule
  \end{tabular}}
\end{table}

\paragraph{Brain-age accuracy on IXI.}
SynthBA achieves MAE~$6.33$ years (RMSE~$7.82$, Pearson $r{=}0.84$,
bias~$-4.15$ years) on the 105-subject IXI subset, outperforming the
calibrated face-age estimate despite using no demographic priors.

\paragraph{Gap correlation.}
On the 93 subjects with both predictions, the face-age gap
($\hat{y}_{\text{face}} - y_{\text{chron}}$) and brain-age gap
($\hat{y}_{\text{brain}} - y_{\text{chron}}$) are positively correlated
(Pearson $r{=}0.31$, Spearman $\rho{=}0.32$, $p{<}0.01$, both).
The moderate effect size suggests a shared aging signal while leaving
substantial independent variance---consistent with face and brain aging
being related but not redundant biomarkers.

\paragraph{Scanner reproducibility on SIMON.}
Across 99 scans on 36 scanners, SynthBA predicted a nearly constant age
(mean $27.4$ years, SD~$1.21$ years) despite chronological age ranging
from 29.6 to 46.4 years (SD~$3.30$ years).  This yields a large
systematic bias (MAE~$16.2$ years, bias~$-16.2$ years), pointing to a
subject-level calibration failure rather than scanner sensitivity.
The very low prediction SD confirms strong scanner-to-scanner reproducibility
of the model output itself.

% -----------------------------------------------------------------------
\section{Discussion and Conclusion}
% -----------------------------------------------------------------------

We presented a proof-of-concept showing that face age and brain age can be
extracted from a single non-defaced T1 MRI.  The key findings are:

\begin{itemize}
  \item MRI-rendered face images yield a detectable chronological age signal
    (MAE~$10.24$ years after calibration, $r{=}0.83$), though less accurate
    than brain-based estimation.
  \item Photorealistic domain transfer via diffusion models \emph{hurts}
    performance, likely due to appearance hallucination conflicting with
    FaceAge's training distribution.
  \item The face-age and brain-age gaps from the same scan are modestly but
    significantly correlated ($r{=}0.31$), suggesting partial shared biology.
  \item SynthBA is highly reproducible across scanners but shows strong
    subject-level bias on SIMON, warranting per-subject calibration.
\end{itemize}

Limitations include the use of a single brain-age model and the lack of
clinical labels.  Future work will extend to multi-contrast renders
(T1/T2/PD), additional brain-age models (SFCN, BrainIAC, MIDIBrainAge),
and datasets with disease labels to test whether face--brain gap discordance
carries diagnostic information.

% Acknowledgments---Will not appear in anonymized version
\midlacknowledgments{We thank the IXI and SIMON data collection teams for
making non-defaced MRI datasets publicly available.}

\bibliography{midl-samplebibliography}

\end{document}
